# Exercises: Inspect Specific GEO Series

In this exercise, you'll retrieve metadata for the following GEO Series and answer:

**GSE IDs:** "GSE201427","GSE166530","GSE162183","GSE186458"

**Questions per GSE**

1.  **What type of data is it? Describe it.** (platform technology; e.g., high-throughput sequencing, array)
2.  **What disease or model is it using?** (infer from series title/summary/platform organism)
3.  **How many samples are there? What type of data is associated with the record?**
4.  **How many control and how many disease?**
5.  **Can you find out any information about the GEO dataset - GSE264108. Why or Why not?**

> Hint: Sample labels are not standardized. Use robust keyword detection (e.g., `control|normal|healthy|wild[- ]?type` vs. `tumou?r|cancer|patient|disease|mutant|treated`). Inspect a few GSM rows if the automatic inference is uncertain.

## Dataset 1 - GSE201427

```{r}
sql <- paste("SELECT * FROM gse WHERE gse.gse = 'GSE201427'")
(rs <- dbGetQuery(con, sql))
```

## Dataset 2 - GSE166530

```{r}
sql <- paste("SELECT * FROM gse WHERE gse.gse = 'GSE166530'")
(rs <- dbGetQuery(con, sql))
```

## Dataset 3 - GSE162183

```{r}
sql <- paste("SELECT * FROM gse WHERE gse.gse = 'GSE162183'")
(rs <- dbGetQuery(con, sql))
```

## Dataset 4 - GSE186458

```{r}
sql <- paste("SELECT * FROM gse WHERE gse.gse = 'GSE186458'")
(rs <- dbGetQuery(con, sql))
```

## Dataset 5 - **GSE264108**

```{r}
sql <- paste("SELECT * FROM gse WHERE gse.gse = 'GSE264108'")
(rs <- dbGetQuery(con, sql))
```
